---
title: "Getting Started with biasProxyR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with biasProxyR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(biasProxyR)
```

## The Problem: Sampling Bias in Data-Poor Regions

When building Species Distribution Models (SDMs), researchers need "background" points to represent the available environment. In data-rich regions, "target-group backgrounds" are used to cancel out sampling bias. 

However, in data-poor regions (like many parts of the Global South), target-group data does not exist. Randomly sampling background points causes the model to learn "where the roads are" rather than "where the species lives." 

## The Solution: `biasProxyR`

`biasProxyR` solves this by using open-source infrastructure data (like OpenStreetMap road networks) to create a "Sampling Effort Proxy" surface. This allows researchers to sample background points that mathematically match the geographic bias of their occurrence data.

## Step 1: Generate the Effort Surface

First, we generate a sampling effort proxy for a country. For this tutorial, we will use Luxembourg because it is small and processes quickly.

```{r generate-effort, message=TRUE, cache=TRUE}`
# Generate a 1000m resolution effort surface for Luxembourg
effort <- generate_effort_surface("Luxembourg", resolution = 1000)

# Plot the surface
terra::plot(effort, main = "Sampling Effort Proxy - Luxembourg", col = terrain.colors(50))
```

*Brighter areas indicate higher probability of human sampling effort (e.g., near cities and major highways).*

## Step 2: Sample Bias-Aware Background Points

Instead of sampling background points randomly, we use the effort surface to weight our sampling. This ensures our background points have the same geographic bias as typical GBIF occurrence data.

```{r sample-bg, message=TRUE, cache=TRUE}`
# Sample 2,000 bias-aware background points
bg_points <- sample_bias_aware_background(effort, n_points = 2000)

# Plot the points over the effort surface
terra::plot(effort, main = "Bias-Aware Background Points", col = terrain.colors(50))
plot(bg_points, add = TRUE, col = "red", pch = 16, cex = 0.5)
```

## Step 3: Diagnose Your Own Data

If you already have occurrence data, you can use the diagnostic tool to see how biased it is before you even run an SDM.

```{r diagnose, message=TRUE, cache=TRUE}`
# Run the diagnostic on our sampled points (pretending they are occurrence data)
diagnostic_results <- diagnose_sampling_bias(bg_points, effort)

# View the calculated bias score
print(diagnostic_results$bias_score_percent)
```

## Next Steps

Now that you have bias-aware background points, you can feed them directly into your favorite SDM package (like `maxnet`, `biomod2`, or `sdm`) alongside your true presence data to build ecologically realistic models!
